---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RiskScorescvd

<!-- badges: start -->
[![R-CMD-check](https://github.com/dvicencio/RiskScorescvd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dvicencio/RiskScorescvd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of RiskScorescvd r package is to calculate the most commonly used cardiovascular risk scores.

We have developed five of the most commonly used risk scores with a dependency (ASCVD [PooledCohort]) making the following available:

* HEART
* EDACS
* GRACE 2.0
* TIMI
* SCORE2/OP
* ASCVD


## Installation

You can install the development version of RiskScorescvd from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dvicencio/RiskScorescvd")
```
## Required variables for each risk score

```{r Plot, echo=FALSE, message=FALSE, fig.width=6, fig.height=8}
library(tidyverse)
index <- c(
  "Age",
  "Gender",
  "diabetes",
  "smoker",
  "hypertension",
  "hyperlipidaemia",
  "family.history",
  "ecg.st.depression",
  "presentation_hstni",
  "typical_symptoms.num",
  "ecg.normal",
  "abn.repolarisation",
  "atherosclerotic.disease",
  "sweating",
  "pain.radiation",
  "pleuritic",
  "palpation",
  "ecg.twi",
  "Δ2nd_hstni",
  "killip.class",
  "systolic.bp",
  "heart.rate",
  "creat",
  "previous.pci",
  "previous.cabg",
  "number.of.episodes.24h",
  "aspirin",
  "total.chol",
  "total.hdl",
  "Ethnicity"
)

# Define function to create heatmap
create_heatmap <- function(index, size, title = FALSE) {
  
  # Create data frame
  df <- data.frame(variable = index)
  
  # Define the variable groupings for each score
  groups <- list(
    HEART = c("typical_symptoms.num", "ecg.normal", "abn.repolarisation", "ecg.st.depression", "Age", "diabetes", "smoker", "hypertension", "hyperlipidaemia", "family.history", "atherosclerotic.disease", "presentation_hstni", "Gender"),
    EDACS = c("Age", "Gender", "diabetes", "smoker", "hypertension", "hyperlipidaemia", "family.history", "sweating", "pain.radiation", "pleuritic", "palpation", "ecg.st.depression", "ecg.twi", "presentation_hstni", "Δ2nd_hstni"),
    GRACE = c("killip.class", "systolic.bp", "heart.rate", "Age", "Gender", "creat", "ecg.st.depression", "presentation_hstni"),
    TIMI = c("Age", "hypertension", "hyperlipidaemia", "family.history", "diabetes", "smoker", "previous.pci", "previous.cabg", "aspirin", "number.of.episodes.24h", "ecg.st.depression", "presentation_hstni", "Gender"),
    ASCVD = c("Gender", "Ethnicity", "Age", "total.chol", "total.hdl", "systolic.bp", "hypertension", "smoker", "diabetes"),
    SCORE2 = c("Age", "Gender", "smoker", "systolic.bp", "Diabetes", "total.chol", "total.hdl")
  )

  # Create new columns with the groupings
  df <- df %>%
    mutate(
      HEART = ifelse(variable %in% groups$HEART, 1, 0),
      EDACS = ifelse(variable %in% groups$EDACS, 1, 0),
      'GRACE 2.0' = ifelse(variable %in% groups$GRACE, 1, 0),
      TIMI = ifelse(variable %in% groups$TIMI, 1, 0),
      ASCVD = ifelse(variable %in% groups$ASCVD, 1, 0),
      'SCORE2/OP' = ifelse(variable %in% groups$SCORE2, 1, 0)
    )
  
  # Reshape the data for plotting
  dt2 <- df %>%
    pivot_longer(cols = HEART:'SCORE2/OP', names_to = "Score", values_to = "Value") %>%
    mutate(
      variable = factor(variable, levels = index),
      Score = factor(Score, levels = c("HEART", "EDACS", "GRACE 2.0", "TIMI", "ASCVD", "SCORE2/OP"))
    )
  
  # Plot the heatmap
  ggplot(dt2, aes(x = Score, y = variable, fill = as.character(Value))) +
    geom_tile(color = "white", linewidth = 1,
              linetype = 1) +
    scale_fill_discrete("Required", labels = c('No', 'Yes')) +
    labs(title = ifelse(title == TRUE, "Heatmap of Variables Used for Each Risk Score", ""),
         x = "Risk Score",
         y = "Variable",
         fill = "Included in Score") +
    scale_y_discrete(labels=c("Age",
                              "Sex",
                              "Diabetes",
                              "Smoker",
                              "Hypertension",
                              "Hyperlipidaemia",
                              "Family History",
                              "ECG ST Depression",
                              "Presentation HSTnI",
                              "Typical Symptoms Num",
                              "ECG Normal",
                              "Abnormal Repolarisation",
                              "Atherosclerotic Disease",
                              "Sweating",
                              "Pain Radiation",
                              "Pleuritic",
                              "Palpation",
                              "ECG TWI",
                              "Δ2nd HSTnI",
                              "Killip Class",
                              "Systolic BP",
                              "Heart Rate",
                              "Creatine",
                              "Previous PCI",
                              "Previous CABG",
                              "Number of Episodes 24h",
                              "Aspirin",
                              "Cholesterol",
                              "HDL",
                              "Ethnicity")) +
    theme(legend.position = "bottom",
          plot.title = element_text(
            hjust = 0.5,
            face = "bold",
            size = (size)
          ),
          legend.title = element_text(
            colour = "black",
            face = "bold",
            size = (size * .833)
          ),
          legend.text = element_text(
            colour = "black",
            size = (size * .833)
          ),
          axis.title = element_text(
            size = (size * .833),
            colour = "black",
            face = "bold"
          ),
          axis.text.x = element_text(colour = "black", face = "bold", size = (size * .6)),
          axis.text.y = element_text(colour = "black", face = "bold", size = (size * .667))) 
    
}
create_heatmap(index, size = 15, title = FALSE)
```


## Example

This is a basic example of how the data set should look to calculate all risk scores available in the package:

```{r example}
library(RiskScorescvd)
# Create a data frame or list with the necessary variables
 # Set the number of rows
 num_rows <- 100
 
 # Create a large dataset with 100 rows
  cohort_xx <- data.frame(
   typical_symptoms.num = as.numeric(sample(0:6, num_rows, replace = TRUE)),
   ecg.normal = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   abn.repolarisation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   ecg.st.depression = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   Age = as.numeric(sample(30:80, num_rows, replace = TRUE)),
   diabetes = sample(c(1, 0), num_rows, replace = TRUE),
   smoker = sample(c(1, 0), num_rows, replace = TRUE),
   hypertension = sample(c(1, 0), num_rows, replace = TRUE),
   hyperlipidaemia = sample(c(1, 0), num_rows, replace = TRUE),
   family.history = sample(c(1, 0), num_rows, replace = TRUE),
   atherosclerotic.disease = sample(c(1, 0), num_rows, replace = TRUE),
   presentation_hstni = as.numeric(sample(10:100, num_rows, replace = TRUE)),
   Gender = sample(c("male", "female"), num_rows, replace = TRUE),
   sweating = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   pain.radiation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   pleuritic = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   palpation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   ecg.twi = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   second_hstni = as.numeric(sample(1:200, num_rows, replace = TRUE)),
   killip.class = as.numeric(sample(1:4, num_rows, replace = TRUE)),
   systolic.bp = as.numeric(sample(40:300, num_rows, replace = TRUE)),
   heart.rate = as.numeric(sample(0:300, num_rows, replace = TRUE)),
   creat = as.numeric(sample(0:4, num_rows, replace = TRUE)),
   cardiac.arrest = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   previous.pci = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   previous.cabg = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   aspirin = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   number.of.episodes.24h = as.numeric(sample(0:20, num_rows, replace = TRUE)),
   total.chol = as.numeric(sample(5:100, num_rows, replace = TRUE)),
   total.hdl = as.numeric(sample(2:5, num_rows, replace = TRUE)),
   Ethnicity = sample(c("white", "black", "asian", "other"), num_rows, replace = TRUE)
 )

  str(cohort_xx)
  
```

## Calculation and Results

This is a basic example of how to calculate all risk scores available in the package and create a new data set with 12 new variables of the calculated and classified risk scores:

```{r example2}
# Call the function with the cohort_xx to calculate all risk scores available in the package
new_data_frame <- calc_scores(data = cohort_xx)
 
# Select columns created after calculation
All_scores <- new_data_frame %>% select(HEART_score, HEART_strat, EDACS_score, EDACS_strat, GRACE_score, GRACE_strat, TIMI_score, TIMI_strat, SCORE2_score, SCORE2_strat, ASCVD_score, ASCVD_strat)

# Observe the results
head(All_scores)

# Create a summary of them to obtain an initial idea of distribution
summary(All_scores)
 
```

