---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RiskScorescvd

<!-- badges: start -->
[![R-CMD-check](https://github.com/dvicencio/RiskScorescvd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dvicencio/RiskScorescvd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of RiskScorescvd r package is to calculate the most commonly used cardiovascular risk scores

We have developed five of the most commonly used risk scores with a dependency (ASCVD [PooledCohort]) making the following available:

* HEART
* EDACS
* GRACE 2.0
* TIMI
* SCORE2/OP
* ASCVD


## Installation

You can install the development version of RiskScorescvd from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dvicencio/RiskScorescvd")
```

## Example

This is a basic example of how the dataset should look to calculate all risk scores available in the package:

```{r example}
library(RiskScorescvd)
# Create a data frame or list with the necessary variables
 # Set the number of rows
 num_rows <- 100
 
 # Create a larger dataset with 100 rows
  cohort_xx <- data.frame(
   typical_symptoms.num = as.numeric(sample(0:6, num_rows, replace = TRUE)),
   ecg.normal = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   abn.repolarisation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   ecg.st.depression = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   Age = as.numeric(sample(30:80, num_rows, replace = TRUE)),
   diabetes = sample(c(1, 0), num_rows, replace = TRUE),
   smoker = sample(c(1, 0), num_rows, replace = TRUE),
   hypertension = sample(c(1, 0), num_rows, replace = TRUE),
   hyperlipidaemia = sample(c(1, 0), num_rows, replace = TRUE),
   family.history = sample(c(1, 0), num_rows, replace = TRUE),
   atherosclerotic.disease = sample(c(1, 0), num_rows, replace = TRUE),
   presentation_hstni = as.numeric(sample(10:100, num_rows, replace = TRUE)),
   Gender = sample(c("male", "female"), num_rows, replace = TRUE),
   sweating = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   pain.radiation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   pleuritic = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   palpation = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   ecg.twi = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   second_hstni = as.numeric(sample(1:200, num_rows, replace = TRUE)),
   killip.class = as.numeric(sample(1:4, num_rows, replace = TRUE)),
   systolic.bp = as.numeric(sample(0:300, num_rows, replace = TRUE)),
   heart.rate = as.numeric(sample(0:300, num_rows, replace = TRUE)),
   creat = as.numeric(sample(0:4, num_rows, replace = TRUE)),
   cardiac.arrest = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   previous.pci = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   previous.cabg = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   aspirin = as.numeric(sample(c(0, 1), num_rows, replace = TRUE)),
   number.of.episodes.24h = as.numeric(sample(0:20, num_rows, replace = TRUE)),
   total.chol = as.numeric(sample(5:100, num_rows, replace = TRUE)),
   total.hdl = as.numeric(sample(2:5, num_rows, replace = TRUE)),
   Ethnicity = sample(c("white", "black", "asian", "other"), num_rows, replace = TRUE)
 )


 # Call the function with the cohort_xx

 new_data_frame <- calc_scores(data = cohort_xx)
 
 summary(new_data_frame)
```
